---
title: "Week 3 Data Wrangling with R - Independent Project"
output: html_notebook
---

```{r}
#### Install tidyverse packages in Rstudio
install.packages("tidyverse")
```

```{r}
#### Load tidyverse for use in our notebook
library(tidyverse)
```
###1. Import Data and load entire dataframe

```{r}
###Import Data
customer_data.df <- read_csv("https://bit.ly/2ZlpzjF")
customer_data.df
```

### 2. Explore dafaframe to understand the data set
```{r}
### Understand the data set by previewing the data frames
head(customer_data.df, 10)
```

```{r}
str(customer_data.df)
```

```{r}
dim(customer_data.df)
```
###3. Standardizing the data Set

```{r}
##Check for duplicate values in the dataframe
customer_data.df[duplicated(customer_data.df), ]
```

##duplicate records were found in the dataset, thus shall be removed

```{r}
##Remove duplicate values in the dataframe, while creating new clean data frame
customer_data.df <- customer_data.df[!duplicated(customer_data.df), ]
```

```{r}
##Confirm deletion of duplicate values in the dataframe
customer_data.df[duplicated(customer_data.df), ]
```
```{r}
##Check missing values in the dataframe
sum(is.na(customer_data.df))
```
##251 records have missing values, thus should be dropped from the dataframe.

```{r}
##Drop records with missing values
customer_data.df <- na.omit(customer_data.df)
```

```{r}
##Confirm missing values
sum(is.na(customer_data.df))
```
```{r}
##Check column names
names(customer_data.df)
```
```{r}
#Change column names to lower case
names(customer_data.df) <- tolower(names(customer_data.df))
names(customer_data.df)
```

```{r}
##Confirm column name case
names(customer_data.df)
```
##Analysis- Determine various churn scenarios
```{r}
##Count total number of churned customers
table(customer_data.df$churn)
```
##Findings: 1863 customers have churned.

```{r}
##Churn analysis by gender
customer_data.df %>%
  filter (churn == 'Yes')%>%
  group_by(gender) %>%
  summarize(Total_Churn = length(churn))
```
##Findings: More Females than males churned service.


```{r}
##Churn analysis by Age Category
customer_data.df %>%
  filter (churn == 'Yes')%>%
  group_by(seniorcitizen) %>%
  summarize(Total_Churn = length(churn))
```
##Findings: Adult population more loyal than youth

